---
title: Subset Every nth Element From a Vector in R
author: Mwavu Kennedy
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
    toc: true
---


## Introduction
I've been using R on a daily basis (literally) precisely for 1 year 6 months 6 days now. One thing I've come to learn is that there are many ways to accomplish most tasks. We as **`useRs`** are usually spoilt for choice.

For some reason about a week ago I needed to subset every second element of a vector. What do I mean?
```{r}
one_hundred <- 1:100
```

I needed to return `2, 4, 6, 8, ...`

Simple enough, right? Right. But what amazed me isn't the simplicity of the question at hand, it was how even as a prolific user of R, the building blocks of the language sometimes slip past me so easily.

Let's go through my thought process:

## Use `seq()`
This was definitely at the top of my mind, no doubts at all. So I wrote something like:
```{r}
indices <- seq(from = 2, to = 100, by = 2)
one_hundred[indices]
```

Pretty good. But... Why do I have to create a vector of indices for that? Why does it feel like I'm typing too much? Okay, this `one_hundred[seq(from = 2, to  = 100, by = 2)]` doesn't make it any better, at least as far as I'm concerned. There must be a better solution.

## The Modulo Operator `%%`
The modulo operator `%%` returns the remainder of the division of one number by another. 
For example, `4 %% 2` would return `0` since:
$$
\frac{4}{2} = 2\;remainder\;0
$$
and `9 %% 7` would return `2` since:
$$ 
\frac{9}{7} = 1\;remainder\;2
$$
Back to our main problem: I basically want all indices which are divisible by $2$ *ie.* Leave a remainder of zero. And this is easy because any number divided by $2$ can only leave a remainder of $1$ or $0$.

Taking our vector:
```{r}
one_hundred %% 2
```

$1$\`s and $0$\`s can be coerced into logicals when subsetting. Values where $1$`s occur will be returned if we do:
```{r}
one_hundred[as.logical(one_hundred %% 2)]
```

To turn the $1$s into $0$s we just have to use the `NOT` operator `!`:
```{r}
one_hundred[!one_hundred %% 2]
```
The `!` operator coerces the numeric `one_hundred %% 2` to a logical vector so we don't have to use `as.logical()` 

Okay, that's a really nice application of my freshman discrete math unit. So at this point I was like:

> "Dude, this is the best solution for this problem. You know I'm kind of an Einstein myself!"

Just to prove to myself that that was the best solution there was I decided to google it. 


## Recycle Shorter Vectors
Vector recycling is one of my all time favorite things about R. The first time I learnt the concept, it was mindblowing. But this time round, it swept me off my feet completely.

The solution was as simple as this:
```{r}
one_hundred[c(FALSE, TRUE)]
```

Tell me that isn't insane! I know it's simple but for some reason I couldn't figure it out on my own. That's one of the best answers on StackOverflow according to me. 

Why? It's:

- Small.
In most cases small never equals good performance but this is one of the few cases where what's advertised is what's delivered.

- Simple.
It's very easy to understand. Straightforward.

- Clean.
It's not obfuscated. 

It's better in every way compared to my initial trials. It was posted by the user `Sven Hohenstein`. [Link to the answer.](https://stackoverflow.com/a/13462110/16246909)

> If logical vectors are used for indexing in R, their values are recycled if the index vector is shorter than the vector containing the values.

## Generalization
Now that we have a the best solution (as far as we know), let's generalize it. 

### $i^{th}$ element
What if you wanted to subset every 9th element? Repeat `TRUE` & `FALSE` the required number of times:
```{r}
# Subset every 9th element:
one_hundred[c(rep(FALSE, times = 8), TRUE)]
```
We repeat `FALSE` $8$ times because we want every $9$th index to be retrieved.

Taking $9$ as our `i`:
```{r}
i <- 9

one_hundred[c(rep(FALSE, times = i - 1), TRUE)]
```

### $(i + n)^{th}$ element
Extending the above logic, let's assume `i` is $9$ and `n` is $10$. So it's like saying we want to subset every $(9 + 10)^{th} = 19^{th}$ element.
```{r}
i <- 9
n <- 10

# So we repeat FALSE "i + n - 1" times:
one_hundred[c(rep(FALSE, times = i + n - 1), TRUE)]
```



## Conclusion
Keep things small. Keep things simple. Keep things clean. Have a single minded focus on the mission. Sounds more like [WhatsApp's engineering culture principles.](https://www.quastor.org/p/how-whatsapp-scaled-to-1-billion)

But at the end of the day it's all a matter of personal preference and what you feel works for you.

Till next time, avoid complicating stuff.


